\myChapter{Theory}\label{functional:theory}
\input{chapters/functional/theory/Types}
\section{\LambdaCalc}
In \partandnameref{Chapter}{background:beginnings}, we briefly sketched the \lambdacalc{}. Now, we shall take the time to do it justice. The fundamentals of the \lambdacalc are simple, unassuming, and somewhat unintuitive. By extending the \lambdacalc{}, we can make it more ``natural'' for reasoning about and even doing programming, but this comes at the cost of reducing its power. It is, however, partly this reduction in power that makes these extensions so appealing.

The rest of this chapter assumes you are familiar with the first-order predicate calculus, in particular the treatment of quantifiers and free and bound variables. If you are not, some of the finer details of the presentation will elude you, but you should still come away with an intuitive understanding of many of the concepts of this chapter.

\input{chapters/functional/theory/PureUntyped}
\input{chapters/functional/theory/Extended}

\section{Bibliographic Notes}
De Bru\ij n indices were first described in~\citet{Bruijn:Lambda:1972}. Other notations for variable binding have been developed more recently; \citet{McBride:Functional:2004}~describes an interesting hybrid that uses de Bru\ij n indices for bound variables and (ideally, meaningful) names for free variables.

We have not made as fine a distinction between the different reduction strategies as that made by~\citet{Sestoft:Demonstrating:2002}. Our description equates call by value with applicative order and call by name with normal order, while he carefully distinguishes these in terms of where reduction can occur.

We have barely scratched the surface of the \lambdacalc and type theory. Those interested in the impact of the \lambdacalc on logic and computer science would find \citet{Barendregt:The-impact:1997} interesting reading. \Citet{Turner:Church:2006} explains the significance of the \lambdacalc and Church's thesis in general for functional programming in but one chapter of a book \citep{Olszewski:Churchs:2006} dedicated to examining Church's thesis seventy years after its postulation. If the theory of the \lambdacalc itself is more in line with your interests, \Citet{Barendregt:The-Lambda:1984}~is the standard reference for the untyped \lambdacalc and develops many aspects of the topic in tremendous detail. The \lambdacalc in one form or another is also often hurriedly introduced as more or less new material where needed in books and articles dealing with functional programming.

The concept of \asword{type} can itself be introduced into the \lambdacalc in a variety of ways and then subsequently elaborated. \citet{Cardelli:On-understanding:1985}~provides a very readable introduction to practical issues of types and programming languages\empause the presentation of types in this chapter is significantly influenced by the presentation therein\empause while \citet{Barendregt:Types:1990}~looks more carefully at the ways typing can be introduced formally into the \lambdacalc{}; our formal definition of the pure, untyped \lambdacalc follows in part the two-page summary given near the beginning of this article. The process we followed of gradually extending the pure untyped \lambdacalc into the typed recursive \lambdacalc with constants follows that of \citet{Hudak:Conception:1989}. \citet{Thompson:Type:1991},~developed from lectures given at the University of Kent and the Federal University of Pernambuco, Recife, Brazil, introduces types in the context of programming languages and constructive logic, while \citet{Pierce:Types:2002}~is a full-length textbook on type theory, and \citet{Pierce:Advanced:2005}~an edited collection of research papers on the topic.
